\documentclass[11pt]{article}

\usepackage[a4paper,left=0.82in,right=0.82in,top=0.85in,bottom=0.9in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{setspace}
\usepackage{longtable}

\hypersetup{colorlinks=true,linkcolor=blue!52!black,citecolor=blue!52!black,urlcolor=blue!52!black}
\setstretch{1.045}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.45em}
\setlist[itemize]{topsep=0.2em,itemsep=0.12em}
\setlist[enumerate]{topsep=0.2em,itemsep=0.12em}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\Lop}{\mathcal L}
\newcommand{\R}{\mathbb R}
\newcommand{\Q}{\mathbb Q}
\newcommand{\1}{\mathbf 1}
\newcommand{\supp}{\operatorname{supp}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\comp}{\operatorname{comp}}
\newcommand{\betti}{\beta}
\newcommand{\eps}{\varepsilon}

\begin{document}

\title{A Colored Tur\'an Proof of the Lollipop Formula}
\author{Siddhartha Mahajan \and Matthias Paulsen \and Paras Chopra}
\date{June 21, 2026}
\maketitle

\begin{abstract}
A lollipop is a circle together with the outward radial ray beginning at one
point of the circle.  Let $a_{\Lop}(n)$ be the maximum number of connected
components of the complement of a union of $n$ lollipops in the plane.  We
prove, for every $n\geq 0$,
\[
 a_{\Lop}(n)=4\binom n2+S(n)+n+1,
\]
where
\[
 S(n)=\max_{\substack{a+b+c+d=n\\0\leq a\leq b\leq c\leq d}}
 \left(3\sum_{1\leq i<j\leq4}m_i m_j-2ab\right),
 \qquad (m_1,m_2,m_3,m_4)=(a,b,c,d).
\]
The upper bound is valid without a general-position assumption.  Its
geometric input is encoded by two graphs: non-close pairs form a $K_4$-free
graph and non-intriguing pairs form a $K_5$-free graph.  A colored Zykov
symmetrization reduces the resulting extremal problem to a weighted blocker
inequality.  The final inequality is an exact $3\times4$ integer-matrix
problem, proved by support compression to star forests.  For the lower bound, we give a polynomial one-parameter family
in which every two distinct nearby copies have exactly four transverse
intersections, and blow up an all-rational perturbation of Karlsson's
four-lollipop configuration.  Every decisive inequality for the base
configuration is checked by exact rational arithmetic; no floating-point
branch or sign decision is used.
\end{abstract}

\section{Statement and notation}\label{sec:statement}

For $c\in\R^2$, $r>0$, and a unit vector $u\in S^1$, write
\[
 C(c,r)=\{x\in\R^2:|x-c|=r\},\qquad
 R(c,r,u)=\{c+tu:t\geq r\}.
\]
The associated \emph{lollipop} is
\[
 L(c,r,u)=C(c,r)\cup R(c,r,u),
\]
and its \emph{anchor} is $a=c+ru$.  No general-position hypothesis is part
of the definition of $a_{\Lop}(n)$.

For $n\geq0$, define
\begin{equation}\label{eq:Sdef}
 S(n)=\max_{\substack{a+b+c+d=n\\0\leq a\leq b\leq c\leq d}}
 \left(3\sum_{1\leq i<j\leq4}m_i m_j-2ab\right),
 \qquad (m_1,m_2,m_3,m_4)=(a,b,c,d).
\end{equation}
Equivalently, for a fixed multiset of four cluster sizes, the penalized
product is the product of the two smallest sizes.

\begin{theorem}\label{thm:main}
For every integer $n\geq0$,
\begin{equation}\label{eq:main}
 a_{\Lop}(n)=4\binom n2+S(n)+n+1.
\end{equation}
\end{theorem}

The proof separates into topology, pair geometry, and an exact colored
extremal theorem.  Cutler-Karlsson-Sloane established the first geometric
bounds and the exact cases through four lollipops; Paulsen introduced the
close-pair and circle-angle obstructions; and Mahajan-Chopra recast those
obstructions as a two-graph problem, determining all cases through $17$ and
also $n=19$ while leaving one-region gaps at $18$ and $20$
\cite{CKS,MahajanChopra,Paulsen}.  Theorem~\ref{thm:main} closes those gaps
and proves the resulting formula for every $n$.

The topological argument below is included in order to remove a frequently
used but unnecessary assertion that every extremizer may simply be assumed
generic.

\section{A region bound valid for arbitrary arrangements}\label{sec:topology}

Let $S^2=\R^2\cup\{\infty\}$ be the one-point compactification.  For a
lollipop $L$, put $\widehat L=L\cup\{\infty\}$.  The compactified stem is a
closed arc from the anchor to $\infty$, so $\widehat L$ is connected and has
first Betti number one.  All homology in this section has coefficients in
$\Q$.

For two lollipops define
\begin{equation}\label{eq:qdef}
 q(L,M)=\betti_0(\widehat L\cap\widehat M)-1.
\end{equation}
The subtraction removes the component containing $\infty$, which is present
for every pair.  Thus $q(L,M)$ counts the other connected components of the
compactified pair intersection, not intersection points with multiplicity.
This definition remains meaningful for tangencies, coincident circles, and
overlapping stems.

\begin{proposition}[topological region inequality]\label{prop:top-region}
For arbitrary lollipops $L_1,\ldots,L_n$, the number $F$ of connected
components of
$\R^2\setminus\bigcup_iL_i$ satisfies
\begin{equation}\label{eq:top-region}
 F\leq n+1+\sum_{1\leq i<j\leq n}q(L_i,L_j).
\end{equation}
If the arrangement has only transverse pairwise crossings, no triple point,
no crossing at an anchor, and no parallel or overlapping stems, then equality
holds and $q(L_i,L_j)$ is the ordinary number of pairwise crossings.
\end{proposition}

\begin{proof}
Realize $S^2$ as the unit sphere in $\R^3$.  Under inverse stereographic
projection, circles, rays, and the point at infinity are semialgebraic; hence
the resulting finite family admits a compatible semialgebraic triangulation.
In particular, every union and intersection below is a compact subcomplex;
its Betti numbers and the number of complement components are finite
\cite{BCR}.  Put
$X_k=\bigcup_{i=1}^k\widehat L_i$.  Since every $\widehat L_i$ contains
$\infty$, each $X_k$ is connected.  The Mayer--Vietoris exact sequence for
$X_{k-1}\cup\widehat L_k$ gives
\[
 \betti_1(X_k)\leq \betti_1(X_{k-1})+\betti_1(\widehat L_k)
 +\betti_0(X_{k-1}\cap\widehat L_k)-1.
\]
As $\betti_1(\widehat L_k)=1$, this becomes
\begin{equation}\label{eq:mvstep}
 \betti_1(X_k)\leq \betti_1(X_{k-1})+
 \betti_0(X_{k-1}\cap\widehat L_k).
\end{equation}
Now
\[
 X_{k-1}\cap\widehat L_k
 =\bigcup_{j<k}(\widehat L_j\cap\widehat L_k).
\]
All sets in this union contain $\infty$.  A union of sets with one common
component has at most one common component plus the sum of their remaining
component counts.  Consequently
\[
 \betti_0(X_{k-1}\cap\widehat L_k)
 \leq1+\sum_{j<k}q(L_j,L_k).
\]
Starting from $\betti_1(X_1)=1$ and iterating \eqref{eq:mvstep} yields
\begin{equation}\label{eq:b1-bound}
 \betti_1(X_n)\leq n+\sum_{i<j}q(L_i,L_j).
\end{equation}
Alexander duality on $S^2$ \cite[Section~3.3]{Hatcher} gives
\[
 F-1=\dim \widetilde H_0(S^2\setminus X_n;\Q)
 =\dim \widetilde H^1(X_n;\Q)=\betti_1(X_n),
\]
which proves \eqref{eq:top-region}.  The case $n=0$ is immediate.

For a generic arrangement, $X_n$ is a finite connected graph.  If $C$ is the
total number of finite crossings, its vertices are the $C$ crossings, the $n$
anchors, and $\infty$, so $V=C+n+1$.  Counting circle and stem segments gives
$E=2C+2n$: each lollipop contributes one baseline circle edge and one baseline
stem edge, and every crossing contributes one subdivision on each of two
lollipops.  Hence
$\betti_1=E-V+1=C+n$, and therefore $F=C+n+1$.  In this case every finite
pair-intersection component is one crossing, so equality also follows in
\eqref{eq:top-region}.
\end{proof}

\section{Pair geometry and the two forced obstructions}\label{sec:geometry}

For two lollipops $L_i=L(c_i,r_i,u_i)$, set $d=c_2-c_1$.
They are called \emph{close} if
\begin{equation}\label{eq:close}
 u_1\cdot u_2\geq0,
\end{equation}
that is, the smaller angle between their stem directions is at most
$\pi/2$.  They are called \emph{intriguing} if either their two circle curves
are disjoint, or
\begin{equation}\label{eq:intriguing}
 |d|^2\leq r_1^2+r_2^2.
\end{equation}
When the circles meet transversely, \eqref{eq:intriguing} says that the angle
between equally oriented circle tangents is at most $\pi/2$.

We first establish all pair bounds, including degenerate cases, directly.

\begin{lemma}[mixed intersections for a close pair]\label{lem:close-mixed}
If $u_1\cdot u_2\geq0$, then
\[
 \#(R(c_1,r_1,u_1)\cap C(c_2,r_2))
 +\#(R(c_2,r_2,u_2)\cap C(c_1,r_1))\leq2.
\]
\end{lemma}

\begin{proof}
A ray meets a circle in at most two points.  Suppose the first mixed
intersection has two distinct points.  Along the supporting line of the first
ray the two intersection parameters have average $d\cdot u_1$.  Both
parameters are at least $r_1$, and they are distinct, so
$d\cdot u_1>r_1$.

If $c_2+t u_2\in C(c_1,r_1)$ for some $t\geq r_2>0$, then
\[
 d\cdot u_1+t(u_2\cdot u_1)
 =(c_2+t u_2-c_1)\cdot u_1
 \leq |c_2+t u_2-c_1|=r_1.
\]
Under \eqref{eq:close} this contradicts $d\cdot u_1>r_1$.  Thus two
intersections in one mixed component force zero in the reverse component.
If neither mixed component has two points, each has at most one, proving the
claim.
\end{proof}

\begin{lemma}[mixed intersections for intriguing circles]\label{lem:intr-mixed}
Assume the circles meet and $|d|^2\leq r_1^2+r_2^2$.
\begin{enumerate}[label=(\roman*)]
\item Each set $R(c_i,r_i,u_i)\cap C(c_j,r_j)$ contains at most one point
outside $C(c_1,r_1)\cap C(c_2,r_2)$.
\item If both mixed sets contain such a point, then the two stems have no
finite intersection.
\end{enumerate}
\end{lemma}

\begin{proof}
Let $p=d\cdot u_1$ and write $|d|^2=p^2+h^2$, where $h$ is the perpendicular
distance from $c_2$ to the first supporting line.  If the first ray meets the
second circle in two distinct points with parameters $t_-<t_+$, then
\[
 t_\pm=p\pm\sqrt{r_2^2-h^2},\qquad t_-\geq r_1.
\]
Put $z=p-r_1$.  We have $z>0$ and
$z\geq\sqrt{r_2^2-h^2}$.  Therefore
\[
 |d|^2-r_1^2-r_2^2
 =2r_1z+z^2-(r_2^2-h^2)>0,
\]
contrary to the hypothesis.  Thus there are not two mixed points.

A singleton mixed intersection outside the common circle intersection cannot
be a tangency.  Indeed, at a tangent parameter $t=p\geq r_1$ one has
$|d|^2=p^2+r_2^2$; the intriguing inequality forces $p\leq r_1$, hence
$p=r_1$, and the tangent point is the first anchor and lies on both circles.
Consequently a singleton outside the common circle intersection is a simple
root beyond the anchor.  Since the quadratic distance function tends to
$+\infty$ and has exactly one root after the anchor, its value at the anchor
is negative.  In other words, the first anchor lies strictly inside the
second disk.  This proves (i), and gives the corresponding interior statement
for either orientation.

Assume now that both outside mixed points exist.  Let
$a_i=c_i+r_i u_i$ be the anchors.  We have
$|a_1-c_2|<r_2$ and $|a_2-c_1|<r_1$.  If the two rays met at a finite point,
write
\[
 x=c_1+(r_1+s)u_1=c_2+(r_2+t)u_2,
 \qquad s,t\geq0.
\]
Because $a_1$ is inside the second disk while every point of the second stem
is outside that disk, $s>0$; similarly $t>0$.  Substituting for $c_2-c_1$ and
using $u_1\cdot u_2\leq1$ gives
\[
 |a_1-c_2|^2=|-s u_1+(r_2+t)u_2|^2
 \geq (r_2+t-s)^2.
\]
The strict inequality $|a_1-c_2|<r_2$ therefore implies $t<s$.  Interchanging
$1$ and $2$ gives $s<t$, a contradiction.  This proves (ii).
\end{proof}

\begin{proposition}[pair savings]\label{prop:pair-savings}
For arbitrary lollipops $L,M$,
\begin{align*}
 q(L,M)&\leq7,\\
 L,M\text{ close}&\Longrightarrow q(L,M)\leq5,\\
 L,M\text{ intriguing}&\Longrightarrow q(L,M)\leq5,\\
 L,M\text{ close and intriguing}&\Longrightarrow q(L,M)\leq4.
\end{align*}
\end{proposition}

\begin{proof}
The circle--circle intersection has at most two connected components (two
points, one point, one coincident circle, or none).  Each mixed ray--circle
intersection has at most two points.  The finite part of a ray--ray
intersection contributes at most one component not connected to $\infty$.
This gives $q\leq2+2+2+1=7$.

For a close pair, Lemma~\ref{lem:close-mixed} reduces the total mixed
contribution to at most two, giving $q\leq2+2+1=5$.

If an intriguing pair has disjoint circles, the circle--circle contribution
vanishes and the general count gives $q\leq0+2+2+1=5$.  Otherwise the circles
meet and satisfy \eqref{eq:intriguing}.  Let $m_{12},m_{21}\in\{0,1\}$ count
mixed points outside the common circle intersection.  Mixed points lying on
both circles do not create additional components beyond the circle--circle
components.  By Lemma~\ref{lem:intr-mixed}, if
$m_{12}=m_{21}=1$ then the finite ray--ray contribution is zero; otherwise it
is at most one.  Hence in either case
\[
 q\leq2+m_{12}+m_{21}+\1_{\{\text{finite ray--ray component}\}}\leq4.
\]
Thus an intersecting intriguing pair actually satisfies the stronger bound
$q\leq4$.  Finally, a pair that is both close and intriguing has bound at
most four in the intersecting case, and at most $0+2+1=3$ in the disjoint
case.
\end{proof}

The two Ramsey-type consequences are next.

\begin{lemma}[forced close pair]\label{lem:forced-close}
Among any four stem directions there is a close pair.
\end{lemma}

\begin{proof}
Place the four directions in cyclic order on $S^1$.  Their four cyclic gaps
sum to $2\pi$, so one gap is at most $\pi/2$.
\end{proof}

\begin{lemma}[forced intriguing pair]\label{lem:forced-intriguing}
Among any five circles there is an intriguing pair.
\end{lemma}

\begin{proof}
Suppose not.  Then every two circle curves meet, and for their center distance
$d_{ij}=|c_i-c_j|$ one has
\begin{equation}\label{eq:nonintr}
 r_i^2+r_j^2<d_{ij}^2\leq(r_i+r_j)^2.
\end{equation}
External tangencies can be removed uniformly: because the family is finite,
there is a number $\lambda>1$, arbitrarily close to one, such that replacing
all $r_i$ by $\lambda r_i$ preserves
$d_{ij}^2>\lambda^2(r_i^2+r_j^2)$ and
$d_{ij}>\lambda|r_i-r_j|$, while making
$d_{ij}<\lambda(r_i+r_j)$.  Thus the enlarged circles meet transversely and
at an obtuse equally oriented angle.  We rename their radii $r_i$.

For each circle set
\[
 v_i=\frac1{r_i}(1,r_i^2-|c_i|^2,c_i)\in\R^4
\]
and equip triples $(\alpha,\beta,x)$ with the symmetric bilinear form
\[
 \langle(\alpha,\beta,x),(\alpha',\beta',x')\rangle
 =\frac12(\alpha\beta'+\beta\alpha')+x\cdot x'.
\]
A direct calculation gives
\[
 \langle v_i,v_i\rangle=1,
 \qquad
 \langle v_i,v_j\rangle
 =\frac{r_i^2+r_j^2-d_{ij}^2}{2r_ir_j}\in(-1,0).
\]
The five vectors are linearly dependent.  In a nonzero relation, positive
and negative coefficients both occur because every first coordinate is
positive.  After changing the sign if necessary, write
\[
 \sum_{i\in P}a_i v_i=\sum_{j\in N}b_j v_j,
 \qquad a_i,b_j>0,
\]
with $1\leq|P|\leq2$.  Put $w=\sum_{i\in P}v_i$.  If $P=\{i\}$, the inner
product of the left side with $w$ is $a_i>0$.  If $P=\{i,k\}$, it is
$(a_i+a_k)(1+\langle v_i,v_k\rangle)>0$.  The inner product of the right side
with $w$ is strictly negative, because every cross inner product is negative.
This contradiction proves the lemma.
\end{proof}

\section{The two-graph reduction}\label{sec:reduction}

Given an arbitrary arrangement, define graphs $D,E$ on the lollipops by
\[
 ij\in D\iff L_i,L_j\text{ are not close},
 \qquad
 ij\in E\iff L_i,L_j\text{ are not intriguing}.
\]
Lemmas~\ref{lem:forced-close} and \ref{lem:forced-intriguing} imply
\begin{equation}\label{eq:forbidden}
 D\text{ is }K_4\text{-free},\qquad E\text{ is }K_5\text{-free}.
\end{equation}
Proposition~\ref{prop:pair-savings} gives, for every pair,
\begin{equation}\label{eq:pointwise-q}
 q(L_i,L_j)\leq
 4+\1_{ij\in D}+\1_{ij\in E}+\1_{ij\in D\cap E}.
\end{equation}
Indeed, the four cases have respective bounds $4,5,5,7$.  Put
\[
 \sigma(D,E)=|D|+|E|+|D\cap E|.
\]
Combining Proposition~\ref{prop:top-region} with \eqref{eq:pointwise-q},
\begin{equation}\label{eq:region-upper-reduction}
 F\leq4\binom n2+\sigma(D,E)+n+1.
\end{equation}
Thus the upper bound in Theorem~\ref{thm:main} follows from the next purely
combinatorial theorem.

\begin{theorem}[colored Tur\'an theorem]\label{thm:colored}
If $D$ is $K_4$-free and $E$ is $K_5$-free on the same $n$-vertex set, then
\begin{equation}\label{eq:colored}
 \sigma(D,E)\leq S(n).
\end{equation}
\end{theorem}

\section{Colored Zykov symmetrization}\label{sec:zykov}

Color each unordered pair by
\[
 0:\text{ neither graph},\quad
 A:E\setminus D,\quad
 B:D\setminus E,\quad
 X:D\cap E.
\]
The weights are $0,1,1,3$, respectively, and their total is $\sigma(D,E)$.

\begin{lemma}[colored Zykov reduction]\label{lem:zykov}
An extremal feasible coloring can be chosen as a blow-up of a complete
nonzero-colored quotient: the vertices are partitioned into classes
$V_1,\ldots,V_k$, pairs inside a class have color $0$, and every pair between
two classes has one fixed color in $\{A,B,X\}$.
\end{lemma}

\begin{proof}
Call two vertices \emph{zero-twins} if they are equal, or their mutual color
is $0$ and their colored neighborhoods agree outside the pair.  This is an
equivalence relation.  Among all feasible colorings of maximum weight, choose
one maximizing the sum of squares of the zero-twin class sizes.

Suppose distinct zero-twin classes $P,Q$ are joined by color $0$.  Then every
$P$--$Q$ pair has color $0$.  Let $d_w(P)$ be the weighted degree of a vertex
of $P$ to vertices outside $P\cup Q$, and define $d_w(Q)$ similarly.  If
$d_w(P)\leq d_w(Q)$, replace every vertex of $P$ by a clone of a representative
of $Q$, keeping all pairs inside $P\cup Q$ color $0$.  The total weight changes
by
\[
 |P|\bigl(d_w(Q)-d_w(P)\bigr)\geq0.
\]
No new $K_4$ in $D$ or $K_5$ in $E$ is created: a new clique containing a
clone from $P$ but no vertex of $Q$ is obtained from an old clique by replacing
the clone by its model in $Q$, while a clique cannot use vertices from both
$P$ and $Q$ because their mutual pairs have color $0$.  If the weighted
degrees are unequal, extremality is contradicted; if they are equal, $P$ and
$Q$ merge and the square-sum tie-breaker is contradicted.  The opposite
inequality is symmetric.  Hence no zero-colored pair joins distinct classes,
and the zero-twin definition makes every cross-class color constant.
\end{proof}

Let the quotient class sizes be $x_1,\ldots,x_k$, with $\sum_i x_i=n$, and
put
\[
 Q=\sum_i x_i^2,
 \qquad P=\sum_{i<j}x_i x_j=\frac{n^2-Q}{2}.
\]
Let $a$ and $b$ be the total weighted masses $\sum x_i x_j$ on quotient
edges of colors $A$ and $B$.  Since all other cross-edges have color $X$,
\begin{equation}\label{eq:sigmaquot}
 \sigma=3P-2a-2b=\frac32(n^2-Q)-2a-2b.
\end{equation}
Because $D$ consists of colors $B,X$, every four quotient vertices contain an
$A$-edge; hence $\alpha(A)\leq3$.  Similarly $\alpha(B)\leq4$.

\section{Weighted blockers}\label{sec:blocker}

For nonnegative integer weights $x_1,\ldots,x_k$ of total $n$, and a graph
$G$ on $[k]$, write
\[
 w(G)=\sum_{ij\in E(G)}x_i x_j.
\]
For $r\geq1$, define
\[
 \rho_r(x)=\min_{\mathcal P}\sum_{P\in\mathcal P}
 \left(\sum_{i\in P}x_i\right)^2,
\]
where $\mathcal P$ ranges over partitions into $r$ possibly empty parts.

\begin{lemma}[weighted Tur\'an theorem]\label{lem:weighted-turan}
If $G$ is $K_{r+1}$-free, then
\[
 w(G)\leq\frac12\bigl(n^2-\rho_r(x)\bigr).
\]
\end{lemma}

\begin{proof}
Zero-weight vertices may be deleted, so assume every $x_i>0$.  Call two
nonadjacent vertices twins when their open neighborhoods agree, and partition
the vertices into twin classes.  Among all maximum-weight
$K_{r+1}$-free graphs, choose one maximizing the sum of the squares of the
total weights of its twin classes.

Suppose two distinct twin classes $P,Q$ are nonadjacent.  Let $d_w(P)$ and
$d_w(Q)$ denote the weighted degrees of their vertices outside
$P\cup Q$; these are constant on each class.  If
$d_w(P)\leq d_w(Q)$, replace every vertex of $P$ by a clone of a
representative of $Q$, keeping all pairs in $P\cup Q$ nonadjacent.  The edge
weight changes by
\[
 \left(\sum_{u\in P}x_u\right)\bigl(d_w(Q)-d_w(P)\bigr)\geq0.
\]
The operation preserves $K_{r+1}$-freeness: a new clique cannot use vertices
from both $P$ and $Q$, and a clique using a clone from $P$ maps to an old
clique by replacing that clone with its model in $Q$.  Strict inequality
contradicts maximality; equality merges $P$ and $Q$ and strictly increases the
tie-breaker.  The opposite ordering is symmetric.  Hence distinct twin
classes are pairwise adjacent.

The graph is therefore complete multipartite, and $K_{r+1}$-freeness leaves
at most $r$ parts.  If their total weights are $s_1,\ldots,s_r$, its edge
weight is
\[
 \sum_{i<j}s_i s_j=\frac12\left(n^2-\sum_i s_i^2\right).
\]
Minimizing the final square-sum over partitions of the original vertex
weights is exactly the definition of $\rho_r(x)$.
\end{proof}

Define
\begin{equation}\label{eq:Mdef}
 M(n)=\min_{\substack{a+b+c+d=n\\a,b,c,d\in\mathbb Z_{\geq0}}}
 \left[\frac32(a^2+b^2+c^2+d^2)+2ab\right].
\end{equation}
For a fixed multiset, the $2ab$ term is minimized by assigning $a,b$ to the
two smallest entries.  Hence
\begin{equation}\label{eq:SM}
 S(n)=\frac32n^2-M(n).
\end{equation}

\begin{lemma}[weighted blocker lemma]\label{lem:blocker}
If $\alpha(A)\leq3$ and $\alpha(B)\leq4$, then
\begin{equation}\label{eq:blocker}
 \frac32Q+2w(A)+2w(B)\geq M(n).
\end{equation}
\end{lemma}

\begin{proof}
The complement of $A$ is $K_4$-free.  Lemma~\ref{lem:weighted-turan} gives
\[
 \frac12(n^2-Q)-w(A)\leq\frac12(n^2-\rho_3(x)),
 \qquad
 w(A)\geq\frac12(\rho_3(x)-Q).
\]
Likewise
$w(B)\geq\frac12(\rho_4(x)-Q)$.  Therefore
\begin{equation}\label{eq:rho}
 \frac32Q+2w(A)+2w(B)
 \geq\rho_3(x)+\rho_4(x)-\frac12Q.
\end{equation}
Choose partitions attaining $\rho_3$ and $\rho_4$, and intersect their parts.
Let $U=(u_{ij})$ be the resulting $3\times4$ nonnegative integer matrix, with
row sums $r_i$ and column sums $c_j$.  Then
\[
 \rho_3=\sum_i r_i^2,
 \qquad \rho_4=\sum_j c_j^2,
 \qquad Q\leq\sum_{i,j}u_{ij}^2,
\]
the last inequality following from $\sum z_\ell^2\leq(\sum z_\ell)^2$ in
each cell.  Thus the right side of \eqref{eq:rho} is at least
\[
 F(U):=\sum_i r_i^2+\sum_jc_j^2-\frac12\sum_{i,j}u_{ij}^2.
\]
The matrix theorem in Section~\ref{sec:matrix} proves $F(U)\geq M(n)$.
\end{proof}

\begin{proof}[Proof of Theorem~\ref{thm:colored}]
Use Lemma~\ref{lem:zykov} and the quotient notation above.  The blocker lemma
implies
$\frac32Q+2a+2b\geq M(n)$.  Equation \eqref{eq:sigmaquot} and
\eqref{eq:SM} give
\[
 \sigma=\frac32n^2-\left(\frac32Q+2a+2b\right)
 \leq\frac32n^2-M(n)=S(n).
\]
\end{proof}

\section{The exact $3\times4$ matrix inequality}\label{sec:matrix}

\begin{theorem}[matrix theorem]\label{thm:matrix}
For every $3\times4$ nonnegative integer matrix $U$ of total sum $n$,
\begin{equation}\label{eq:matrix}
 F(U)=\sum_i r_i^2+\sum_jc_j^2-\frac12\sum_{i,j}u_{ij}^2\geq M(n).
\end{equation}
\end{theorem}

View positive cells as weighted edges of $K_{3,4}$.  Expanding row and column
squares gives
\begin{equation}\label{eq:Fedge}
 F(U)=\frac32\sum_eu_e^2+2\sum_{e\sim f}u_eu_f,
\end{equation}
where adjacency means sharing a row or column vertex.

\begin{lemma}[support compression]\label{lem:compression}
Every nonnegative integer matrix $U$ can be replaced by a matrix $U'$ of the
same total sum, with $F(U')\leq F(U)$, whose support graph is a star forest.
\end{lemma}

\begin{proof}
If the support contains a cycle, add $\eps$ and $-\eps$ alternately around it.
All row and column sums remain fixed, while
$-\frac12\sum u_{ij}^2$ is a concave quadratic in $\eps$.  The allowed
integer values form a finite interval containing zero.  At least one endpoint
has value no larger than at zero and makes an entry vanish.  Repetition removes
all cycles.

Suppose next that a tree component has diameter at least four.  On four
consecutive path edges of masses $x_1,x_2,x_3,x_4$, use
\[
 (x_1,x_2,x_3,x_4)\mapsto
 (x_1+\eps,x_2-\eps,x_3+\eps,x_4-\eps).
\]
The three internal vertex sums are fixed and the two endpoint-square terms
contribute $2\eps^2$.  The four cell-square terms contribute
$-\frac12\cdot4\eps^2=-2\eps^2$.  Hence $F$ is affine in $\eps$ on the
allowed integer interval.  An endpoint does not increase $F$ and deletes an
edge.  Repetition leaves a forest of diameter at most three.

A diameter-three component is a double star.  Let its central edge have mass
$y$.  Choose leaf masses $x,z$ on opposite sides, and let $A,B$ be the sums of
the other leaf masses at the respective centers.  Deleting the central edge
and adding $y$ to the $x$-edge gives, by \eqref{eq:Fedge},
\[
 F_{\rm old}-F_x=y(2B+2z-x).
\]
Adding $y$ to the $z$-edge instead gives
\[
 F_{\rm old}-F_z=y(2A+2x-z).
\]
Both expressions cannot be negative, since that would imply
$x>2B+2z\geq2z$ and $z>2A+2x\geq2x$.  One move therefore does not increase
$F$ and deletes the central edge.  Repetition removes every diameter-three
component.  A forest of diameter at most two is a star forest.
\end{proof}

For a star of degree $t$, edge masses $z_1,\ldots,z_t$, and total mass
$s=\sum z_i$, its contribution is
\begin{equation}\label{eq:star}
 s^2+\frac12\sum_i z_i^2
 \geq\left(1+\frac1{2t}\right)s^2
 =\frac{s^2}{\eta_t},
 \qquad \eta_t=\frac{2t}{2t+1}.
\end{equation}

\begin{lemma}[star-forest bound]\label{lem:starforest}
Every star-forest matrix $U$ of total mass $n$ satisfies $F(U)\geq M(n)$.
\end{lemma}

\begin{proof}
If $n=0$, then $U=0$ and $F(U)=M(0)=0$.  Assume henceforth that $n>0$.
Let the nontrivial star components have degrees $t_\nu$ and masses $s_\nu$.
Except for degree list $(2,1,1)$,
\begin{equation}\label{eq:Gamma}
 \Gamma:=\sum_\nu\eta_{t_\nu}\leq2.
\end{equation}
Indeed, there are at most three nontrivial components because each uses a row
vertex.  One or two components give $\Gamma<2$.  With three components, their
vertex counts satisfy $\sum(t_\nu+1)\leq7$, so the only degree lists are
$(1,1,1)$ and $(2,1,1)$; the former has $\Gamma=2$.

For all nonexceptional degree lists, \eqref{eq:star} and Cauchy's inequality
give
\[
 F(U)\geq\sum_\nu\frac{s_\nu^2}{\eta_{t_\nu}}
 \geq\frac{n^2}{\Gamma}\geq\frac{n^2}{2}.
\]
For $n\geq4$, $M(n)\leq n^2/2$: if $n=4q+r$, use respectively
\[
 (q,q,q,q),\ (q,q,q,q+1),\ (q,q,q+1,q+1),\ (q,q+1,q+1,q+1)
\]
for $r=0,1,2,3$ in \eqref{eq:Mdef}; direct substitution gives the claimed
inequality.  For $1\leq n\leq3$, the inequality $z^2\geq z$ for
$z\in\mathbb Z_{\geq0}$ gives $M(n)\geq3n/2$, and the quadruples displayed
in Section~\ref{sec:values} attain equality.  Meanwhile
\[
 F(U)=\frac32n+3\sum_{i,j}\binom{u_{ij}}2
 +2\sum_{e\sim f}u_eu_f\geq\frac32n.
\]

It remains to treat degree list $(2,1,1)$.  The degree-two star must be centered
at a row: if centered at a column it uses two rows, leaving only one row for
two disjoint edges.  After row and column permutations the positive entries
are
\[
 \begin{pmatrix}a&b&0&0\\0&0&c&0\\0&0&0&d\end{pmatrix}.
\]
Then
\[
 F(U)=\frac32(a^2+b^2+c^2+d^2)+2ab\geq M(n)
\]
by the definition of $M(n)$.
\end{proof}

\begin{proof}[Proof of Theorem~\ref{thm:matrix}]
Compress $U$ to a star-forest matrix $U'$ by Lemma~\ref{lem:compression}.
Then $F(U)\geq F(U')\geq M(n)$ by Lemma~\ref{lem:starforest}.
\end{proof}

Combining Theorem~\ref{thm:colored} with
\eqref{eq:region-upper-reduction} proves the upper bound in
Theorem~\ref{thm:main} for all arrangements, including degenerate ones.

\section{The lower construction}\label{sec:lower}

We now construct generic arrangements attaining the upper bound.  Two
stability lemmas are stated explicitly because stems are unbounded.

\subsection{An all-rational four-lollipop base}

Write $Q_{\rm anch}(a,r,u)$ for the lollipop specified by its anchor
$a\in\Q^2$, radius $r\in\Q_{>0}$, and unit stem direction $u\in\Q^2$, and
write $Q_{\rm cent}(c,r,u)$ when the center is specified instead.  Set
\[
 u_0=(1,0),\qquad
 u_1=\frac1{145}(144,-17),\qquad
 u_2=\frac1{85}(-36,-77),\qquad
 u_3=\frac1{145}(-17,144).
\]
The identities $144^2+17^2=145^2$ and $36^2+77^2=85^2$ show that every
$u_i$ is a unit vector.  Consider
\begin{align*}
 Q_0&=Q_{\rm anch}\bigl((0,0),200,u_0\bigr),\\
 Q_1&=Q_{\rm cent}\bigl((9/20,2/5),11/20,u_1\bigr),\\
 Q_2&=Q_{\rm anch}\bigl((23/20,13/20),100,u_2\bigr),\\
 Q_3&=Q_{\rm anch}\bigl((21/20,-1/20),100,u_3\bigr).
\end{align*}
This is a rational perturbation of Karlsson's four-lollipop construction.

\begin{proposition}[exact rational base table]\label{prop:base}
Every component intersection in the following table is transverse; the
supporting stem lines are pairwise nonparallel; no listed intersection is an
anchor; and the component counts are
\[
\begin{array}{c|cccc|c}
\toprule
\text{pair}&C_i\cap C_j&R_i\cap C_j&R_j\cap C_i&R_i\cap R_j&\text{total}\\
\midrule
Q_0,Q_1&2&2&0&1&5\\
Q_0,Q_2&2&2&2&1&7\\
Q_0,Q_3&2&2&2&1&7\\
Q_1,Q_2&2&2&2&1&7\\
Q_1,Q_3&2&2&2&1&7\\
Q_2,Q_3&2&2&2&1&7\\
\bottomrule
\end{array}
\]
Thus the base has forty crossings.
\end{proposition}

\begin{proof}
All coordinates are rational, so the verification reduces to strict rational
inequalities.  For an ordered pair $i,j$, put $d=c_j-c_i$,
$p=d\cdot u_i$, and
\[
 \Delta_{ij}=r_j^2-(|d|^2-p^2),\qquad
 A_{ij}=|d-r_i u_i|^2-r_j^2,\qquad
 T_{ij}=p-r_i.
\]
The circle--circle component has two transverse points exactly when
\[
 (r_i-r_j)^2<|d|^2<(r_i+r_j)^2.
\]
When $\Delta_{ij}>0$, the supporting line meets the circle twice.  The
accepted ray component has one point if $A_{ij}<0$, two points if
$A_{ij}>0$ and $T_{ij}>0$, and no point if $A_{ij}>0$ and $T_{ij}<0$.
These alternatives follow by evaluating the convex quadratic
$|c_i+t u_i-c_j|^2-r_j^2$ at $t=r_i$ and locating its vertex at $t=p$.

For the two supporting stem lines, put
$D_{ij}=\det(u_i,u_j)$.  When $D_{ij}\ne0$, the parameters of their unique
line intersection are
\[
 t_i=\frac{\det(d,u_j)}{D_{ij}},\qquad
 t_j=\frac{\det(d,u_i)}{D_{ij}}.
\]
It is a ray--ray point precisely when $t_i>r_i$ and $t_j>r_j$.

Substitution of the displayed rational data into these tests gives the six
rows of the table.  The ancillary program
\path{scripts/certify_rational_base.py} performs exactly those
substitutions using \texttt{fractions.Fraction}; its branches and sign
decisions use no floating-point arithmetic.  The full reduced fractions for
every decisive expression are recorded in
\path{verification/rational_base_verification.json}.  In particular all
circle inequalities are strict, every accepted mixed root is simple and away
from its anchor, all $D_{ij}$ are nonzero, and both ray--ray parameters are
strictly beyond their anchor thresholds.  The smallest absolute decisive
quantity in this exact calculation is $151/23400>0$.
\end{proof}

\subsection{Pairwise chambers and a polynomial local family}

\begin{lemma}[pairwise chamber stability]\label{lem:chamber}
Let $L,M$ be a pair for which the circles are neither tangent nor coincident,
every accepted circle--stem intersection is transverse and away from an
anchor, the supporting stem lines are nonparallel, and their finite line
intersection is not an anchor.  Then there are parameter neighborhoods of
$L$ and $M$ on which the number of intersections in each of the four
component pairs is constant.  All the stated strict conditions remain valid
in those neighborhoods.
\end{lemma}

\begin{proof}
Circle--circle alternatives are determined by strict inequalities involving
the center distance and the sum and difference of radii.  A mixed component
is determined by a quadratic in the stem parameter.  Simple accepted roots
vary continuously.  A double root can occur only outside the accepted
half-line under the hypotheses; its positive distance from the anchor
threshold persists under a small perturbation.  Thus the number of accepted
roots is locally constant.  For the stems, the two supporting-line parameters
are rational functions of the geometric parameters, with denominator the
nonzero determinant of the direction vectors.  Their strict positions
relative to the two anchor thresholds persist.  Intersecting the finitely
many resulting open conditions proves the lemma.
\end{proof}

The following family replaces a previously proposed motion that produced
only three intersections.

\begin{lemma}[local four-crossing family]\label{lem:local4}
Let $\Lambda$ be the unit lollipop centered at the origin with stem
$\{(q,0):q\geq1\}$.  There is a continuous family
$(\Gamma_t)_{0\leq t\leq1/4}$, with $\Gamma_0=\Lambda$, such that every two
distinct members have exactly four transverse crossings, none at an anchor.
\end{lemma}

\begin{proof}
Set
\[
 c_t=(3t,3t),\qquad v_t=(1-t^2,-2t),\qquad r_t=1+t^2,
\]
and let $\Gamma_t$ have center $c_t$, radius $r_t$, and stem
$\{c_t+qv_t:q\geq1\}$.  Since
$|v_t|^2=(1-t^2)^2+4t^2=(1+t^2)^2=r_t^2$, the stem is radial and outward.

Fix $0\leq s<t\leq1/4$ and put $\delta=t-s$.  The squared center distance is
$18\delta^2$, while the radius difference is $\delta(s+t)$.  Hence
\[
 \delta^2(s+t)^2<18\delta^2<(2+s^2+t^2)^2,
\]
so the circles meet transversely twice.

For the stem of $\Gamma_s$ against the circle of $\Gamma_t$, put
\[
 F_{s,t}(q)=|c_s+qv_s-c_t|^2-r_t^2.
\]
A direct expansion gives
\[
 F_{s,t}(1)=-\delta B_{s,t},
\]
where
\[
 B_{s,t}=6+8s-16t-6s^2+s^3+s^2t+st^2+t^3
 \geq6-16t-6s^2\geq\frac{13}{8}.
\]
Moreover
\[
 \frac12F'_{s,t}(1)
 =(1+s^2)^2-3\delta(1-2s-s^2)
 \geq1-3\delta\geq\frac14.
\]
Since $F''_{s,t}=2|v_s|^2>0$, the function is strictly increasing on
$[1,\infty)$, begins negative, and tends to infinity.  It has exactly one
simple zero there.

In the reverse direction, define
\[
 G_{s,t}(q)=|c_t+qv_t-c_s|^2-r_s^2.
\]
Then
\[
 G_{s,t}(1)=\delta C_{s,t},
\quad
 C_{s,t}=6-16s+8t-6t^2+s^3+s^2t+st^2+t^3
 >6-8t-6t^2\geq\frac{29}{8},
\]
and
\[
 \frac12G'_{s,t}(1)
 =(1+t^2)^2+3\delta(1-2t-t^2)>0.
\]
Thus $G_{s,t}$ is positive and increasing on $[1,\infty)$, so this mixed
component is empty.

Finally,
\[
 \det(v_s,v_t)=2(s-t)(1+st)\ne0.
\]
Solving $c_s+\lambda v_s=c_t+\mu v_t$ gives
\[
 \lambda=\frac{3(1+2t-t^2)}{2(1+st)},\qquad
 \mu=\frac{3(1+2s-s^2)}{2(1+st)}.
\]
Both exceed one, because
\[
 1+6t-3t^2-2st>0,
 \qquad 1+6s-3s^2-2st>0;
\]
the negative terms in either expression have total magnitude at most
$5/16$.  Hence the stems meet once beyond both anchors.  A point strictly
beyond an anchor lies outside its own circle, which also prevents coincidence
of this point with any circle crossing and prevents the unique mixed point
from being a triple component intersection.  The total is
$2+1+0+1=4$.
\end{proof}

\begin{lemma}[genericization inside a chamber]\label{lem:genericize}
Let a finite lollipop arrangement have transverse pairwise crossings away from
anchors and pairwise nonparallel supporting stem lines.  Every sufficiently
small product neighborhood preserving its pairwise chambers contains an
arrangement with no finite triple point and with all pairwise component counts
unchanged.
\end{lemma}

\begin{proof}
Shrink to a product of connected coordinate balls on which
Lemma~\ref{lem:chamber} preserves every pairwise count and strict condition.
Choose the lollipops successively.  After $k-1$ have been chosen generically,
their finite pairwise intersection set $P$ is finite.  For a fixed $p\in P$,
the parameters $(c,r,\theta)$ for which the next circle contains $p$ lie in
the zero set of the nonzero real-analytic function
\[
 g_p(c,r,\theta)=|p-c|^2-r^2.
\]
The parameters for which its supporting stem line contains $p$ lie in the
zero set of the nonzero real-analytic function
\[
 h_p(c,r,\theta)=\det(p-c,u_\theta).
\]
Neither function vanishes identically on a coordinate ball: vary $r$ for
$g_p$, and vary $c$ transversely to $u_\theta$ for $h_p$.  A proper
real-analytic zero set is closed with empty interior, hence nowhere dense.
A finite union of these closed nowhere-dense sets cannot contain the open
parameter ball.  Choose the $k$-th lollipop outside that union.  It avoids all
old pairwise intersection points, while the chamber conditions retain all
pairwise counts, transversality, nonparallelity, and anchor exclusions.
Induction removes every finite triple point.
\end{proof}

\begin{lemma}[four-cluster blow-up]\label{lem:blowup}
For any nonnegative integers $m_0,m_1,m_2,m_3$, there is a generic arrangement
of $n=\sum_i m_i$ lollipops having
\begin{equation}\label{eq:Lm}
 4\sum_i\binom{m_i}{2}+5m_0m_1
 +7\sum_{\substack{i<j\\(i,j)\ne(0,1)}}m_i m_j
\end{equation}
finite crossings.
\end{lemma}

\begin{proof}
By Proposition~\ref{prop:base} and Lemma~\ref{lem:chamber}, choose
neighborhoods $\mathcal U_i$ of $Q_i$ such that every cross-cluster pair in
$\mathcal U_i\times\mathcal U_j$ has the base component counts and nonparallel
stems.  Map the local family of Lemma~\ref{lem:local4} to $Q_i$ by a
similarity.  A sufficiently short initial interval lies in $\mathcal U_i$.
Choose $m_i$ distinct parameters from it.  Same-cluster pairs have four
crossings, while cross-cluster pairs have five for $(0,1)$ and seven otherwise.
All pairwise intersections are transverse and away from anchors.  Apply
Lemma~\ref{lem:genericize} in a sufficiently small product chamber to remove
accidental triple points without changing any pair count.  Summing gives
\eqref{eq:Lm}.
\end{proof}

For $n=\sum_i m_i$, subtracting $4\binom n2$ from \eqref{eq:Lm} gives
\[
 3\sum_{i<j}m_i m_j-2m_0m_1.
\]
For a fixed multiset of cluster sizes, this is maximized by assigning the two
smallest sizes to $m_0,m_1$.  Maximizing over all quadruples gives $S(n)$.
The final arrangement is generic, so the equality case of
Proposition~\ref{prop:top-region} gives
\[
 a_{\Lop}(n)\geq4\binom n2+S(n)+n+1.
\]
Together with the upper bound already proved, this completes the proof of
Theorem~\ref{thm:main}.

\section{Asymptotics and values}\label{sec:values}

The continuous relaxation of \eqref{eq:Mdef} has its unique minimum at
\[
 a=b=\frac{3n}{16},\qquad c=d=\frac{5n}{16}.
\]
Rounding changes the quadratic objective by $O(1)$, so
\[
 M(n)=\frac{15}{32}n^2+O(1),
 \qquad S(n)=\frac{33}{32}n^2+O(1),
\]
and therefore
\[
 a_{\Lop}(n)=\frac{97}{32}n^2-n+O(1).
\]
The first values are shown below.

\begin{center}
\begin{tabular}{@{}rcrr@{}}
\toprule
$n$&one optimal $(a,b,c,d)$&$S(n)$&$a_{\Lop}(n)$\\
\midrule
0&(0,0,0,0)&0&1\\
1&(0,0,0,1)&0&2\\
2&(0,0,1,1)&3&10\\
3&(0,1,1,1)&9&25\\
4&(1,1,1,1)&16&45\\
5&(1,1,1,2)&25&71\\
6&(1,1,2,2)&37&104\\
7&(1,2,2,2)&50&142\\
8&(1,2,2,3)&65&186\\
9&(1,2,3,3)&83&237\\
10&(2,2,3,3)&103&294\\
11&(2,2,3,4)&124&356\\
12&(2,2,4,4)&148&425\\
13&(2,3,4,4)&174&500\\
14&(2,3,4,5)&201&580\\
15&(2,3,5,5)&231&667\\
16&(3,3,5,5)&264&761\\
17&(3,3,5,6)&297&859\\
18&(3,3,6,6)&333&964\\
19&(3,4,6,6)&372&1076\\
20&(4,4,6,6)&412&1193\\
\bottomrule
\end{tabular}
\end{center}

\section*{Reproducibility statement}

The ancillary bundle contains the complete \LaTeX{} source, the exact
rational verifier for Proposition~\ref{prop:base}, an independent finite
checker for the colored and matrix extremal problems, their machine outputs,
and SHA-256 checksums.  The base verifier uses no floating-point branch or sign
decision.  The finite extremal checks are corroborative only; the proofs in
Sections~\ref{sec:zykov}--\ref{sec:matrix} are symbolic and do not depend on
them.

\begin{thebibliography}{9}

\bibitem{CKS}
D. O. H. Cutler, J. Karlsson, and N. J. A. Sloane,
\emph{Cutting a Pancake with an Exotic Knife},
arXiv:2511.15864v3, 2026.

\bibitem{MahajanChopra}
S. Mahajan and P. Chopra,
\emph{A Two-Graph Refinement of Paulsen's Lollipop Bounds},
arXiv:2606.06064, 2026.

\bibitem{OEIS}
OEIS Foundation Inc.,
\emph{The On-Line Encyclopedia of Integer Sequences}, entry A389624,
2026.

\bibitem{Paulsen}
M. Paulsen,
\emph{The Lollipop Problem}, 2026,
\url{https://math.mp42.de/lollipop/}.


\bibitem{BCR}
J. Bochnak, M. Coste, and M.-F. Roy,
\emph{Real Algebraic Geometry}, Springer, 1998.

\bibitem{Hatcher}
A. Hatcher,
\emph{Algebraic Topology}, Cambridge University Press, 2002.

\bibitem{Turan}
P. Tur\'an,
On an extremal problem in graph theory,
\emph{Matematikai es Fizikai Lapok} \textbf{48} (1941), 436--452.

\bibitem{Zykov}
A. A. Zykov,
On some properties of linear complexes,
\emph{Matematicheskii Sbornik} \textbf{24} (1949), 163--188.

\end{thebibliography}

\end{document}
